---
title: "01b_cusip_universe"
author: "Vinci Chan"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc_expand: true
    embed-resources: true
execute:
  warning: false
  message: false
---

Points to note:

1. Quarters unavailable for download: 1997Q4
2. 2002Q3 and 2002Q4 has fewer than normal CUSIPS which can affect downstream extraction

# 1. Load Package

```{r}
library(tidyverse)
library(DBI)
library(RSQLite)
library(pdftools)

source("R/config.R")
source("R/utils_dates.R")
source("R/utils_db.R")
```

# 2. Setup

```{r}
dates <- make_quarter_tbl(
  start_year    = 1996, # start at 1996 as the lists are only made available since 1996
  end_year      = cfg$end_year,
)

qa_dir_01b <- file.path(cfg$qa_dir, "01b")
dir.create(qa_dir_01b, recursive = TRUE, showWarnings = FALSE)
```

# 3. Create SEC list pdf urls

```{r}
input <- dates %>% 
  mutate(
    link = case_when(
      year == 2004 & qtr == 1 ~ 
        "https://www.sec.gov/divisions/investment/13f-list.pdf", # special case
      year < 2021 ~ 
        paste0("https://www.sec.gov/divisions/investment/13f/13flist", year, "q", qtr, ".pdf"),
      year == 2021 & qtr <= 2 ~
        paste0("https://www.sec.gov/divisions/investment/13f/13flist", year, "q", qtr, ".pdf"),
      TRUE ~
        paste0("https://www.sec.gov/files/investment/13flist", year, "q", qtr, ".pdf")
    )
  )
```

# 3. Functions

## 3.1. Parse one page of the pdf into a table

```{r}
parse.one.page <- function(page, mode = c("post", "pre")) {
  mode <- match.arg(mode)

  if (!any(grepl("CUSIP NO", page))) return(NULL)

  # remove rows above header
  page <- page[min(which(grepl("CUSIP NO", page))):length(page)]

  if (mode == "post") {
    header <- page[1]

    seps <- c(
      0,
      gregexpr("ISSUER NAME", header)[[1]][1] - 1,
      gregexpr("ISSUER DESCRIPTION", header)[[1]][1] - 1,
      gregexpr("STATUS", header)[[1]][1] - 1,
      max(nchar(page))
    )

    split.row <- function(row, seps) {
      n_cols <- length(seps) - 1
      out <- character(n_cols)

      for (i in seq_len(n_cols)) {
        start <- seps[i] + 1
        end   <- seps[i + 1]
        out[i] <- trimws(substr(row, start, end))
      }
      out
    }

    rows <- lapply(page, split.row, seps = seps)
    cn   <- c("cusip", "issuer_name", "issuer_description", "status")

  } else { # mode == "pre"
    seps <- c(0, 10, max(nchar(page)))

    split.row <- function(row, seps) {
      vapply(
        seq_len(length(seps) - 1),
        function(i) trimws(substr(row, seps[i] + 1, seps[i + 1])),
        character(1)
      )
    }

    rows <- lapply(page, split.row, seps = seps)
    cn   <- c("cusipno", "other")
  }

  matrix <- do.call(rbind, rows)
  if (nrow(matrix) <= 2) return(NULL)

  colnames(matrix) <- cn
  as_tibble(matrix[-1, , drop = FALSE])
}
```

## 3.2. Download individual pdfs, calls parse.one.page() and create a table for the whole pdf

```{r}
download.sec.list <- function(link, year) {
  pre2004 <- year < 2004
  mode    <- if (pre2004) "pre" else "post"

  tmp_pdf <- file.path(cfg$staging_dir, "tmp_cusip.pdf")

  ok <- FALSE
  for (attempt in 1:3) {
    cat(sprintf("Attempt %d/3 for %s\n", attempt, basename(link)))
    res <- try(
      download.file(
        link,
        tmp_pdf,
        headers = c("User-Agent" = cfg$sec_user_agent),
        mode    = "wb",
        quiet   = TRUE
      ),
      silent = TRUE
    )

    if (!inherits(res, "try-error")) {
      ok <- TRUE
      break
    }

    cat("Download failed, waiting 30s before retry \n")
    Sys.sleep(30)
  }

  if (!ok) {
    unlink(tmp_pdf)
    stop("Failed to download PDF after 3 attempts ", link)
  }

  if (pre2004) {
    pdf   <- pdf_ocr_text(tmp_pdf)
  } else {
    pdf   <- pdf_text(tmp_pdf)
  }

  pages       <- strsplit(pdf, "\n")
  page_tables <- lapply(pages, parse.one.page, mode = mode)

  unlink(tmp_pdf)

  out <- page_tables %>% purrr::compact() %>% bind_rows()
  as_tibble(out)
}
```

## 3.3. Clean table

### 3.3.1. Post-2004

```{r}
clean.cusip.list.post <- function(cusip_df) {
  cusip_df %>% 
    filter(!is.na(cusip)) %>% 
    mutate(
      asterisk = grepl("\\*", cusip),
      cusip = gsub("\\s|\\*", "", cusip)
    ) %>% 
    relocate(asterisk, .after = cusip) %>% 
    mutate(nchar = nchar(cusip)) %>% 
    filter(nchar == 9) %>% 
    distinct(cusip, .keep_all = TRUE)
}
```

### 3.3.2. Pre-2004

```{r}
checksum.check <- function(cusip, return.check = TRUE) {
  # helper: contribution of one char at position pos
  checksum.value <- function(x, pos) {
    # map digit → 0–9, letter → 10–35
    x <- ifelse(grepl("[[:digit:]]", x),
                as.numeric(x),
                which(x == LETTERS) + 9)

    # double if even position
    x <- ifelse(pos %% 2 == 0, 2 * x, x)

    # sum decimal digits (e.g. 16 → 1 + 6 = 7)
    x <- floor(x / 10) + (x %% 10)
    x
  }

  # 1. remove spaces
  cusip <- gsub("\\s", "", cusip)

  # 2. split each CUSIP into characters
  cusip_split <- strsplit(cusip, "")

  # 3. compute sum of contributions over positions 1–8
  sums <- lapply(cusip_split, function(chars) {
    total <- 0
    for (pos in 1:8) {
      total <- total + checksum.value(chars[pos], pos)
    }
    total
  })

  # 4. implied check digit for each CUSIP
  value <- lapply(sums, function(x) (10 - (x %% 10)) %% 10)

  # 5. compare to 9th character
  check <- lapply(seq_along(value), function(i) {
    chars9 <- cusip_split[[i]][9]
    if (is.na(chars9)) return(FALSE)
    chars9 == as.character(value[[i]])
  })
  check <- lapply(check, function(x) ifelse(is.na(x), FALSE, x))

  # 6) return either logical flags or check digits
  if (return.check) {
    unlist(check)
  } else {
    unlist(value)
  }
}
```

```{r}
clean.cusip.list.pre <- function(cusip_df) {
  table <- cusip_df
  table <- table %>%
    filter(!is.na(cusipno), !is.na(other)) %>% # remove NAs
    filter(!cusipno == "", !other == "") %>% # remove empty lines
    filter(!grepl("total", cusipno, ignore.case = TRUE)) # remove cusipnos that have "TOTAL"

  cusips <- trimws(toupper(table$cusipno))
  others <- trimws(table$other)

  # list of OCR issues, their replacements, and other erros
  issues <- c('4%', '%', '\\.', '^\\|', '\\|', '8\\&$', '\\&', 'J/', '/ ', '/', ')', 'G@|4@|@', ' 1] ', '1]|]j|]', '¥Y|¥', '§', '\\?', '\\$', '}$', '}')
  replace <- c('', '4', '', '', '1', '', '8', 'J', 'J', '', 'J', '4', '11', '1', 'Y', '4', '7', '3', '1', '')
  errors <- c(';', "'", '\\+', '°', '=', ',', '\\(', '~', '¢', '\\_', '\\:', '\\#', '\\.', '-', '—')

  # remove errors
  for (i in seq_len(length(errors))) {
    cusips <- gsub(errors[i], "", cusips)
  }

  # validate cusips
  check <- checksum.check(cusips)

  # for cusips that returned FALSE in checksum
  for (i in seq_len(length(issues))) {
    candidate1 <- trimws(gsub(issues[i], replace[i], cusips[check != TRUE])) # correct issues with issues and replace
    candidate2 <- trimws(gsub("[^[:alnum:][:space:]]", "", cusips[check != TRUE])) # remove non-alphanumeric and non-space characters

    candidate1_check <- checksum.check(candidate1)
    candidate2_check <- checksum.check(candidate2)

    # default to candidate 1
    combined <- candidate1
    combined_check <- candidate1_check
    inds <- candidate2_check == TRUE & !is.na(candidate2_check)
    combined[inds] <- candidate2[inds]
    combined_check[inds] <- candidate2_check[inds]

    cusips[check != TRUE] <- combined
    check[check!= TRUE] <- combined_check
  }

  cusips <- gsub("[^[:alnum:]]", "", cusips) # remove all spaces

  # lengthen cusips shorter than 9 characters
  short <- nchar(cusips) < 9
  cusips[short] <- paste0(cusips[short], substr(gsub("[^[:alnum:]]", "", others[short]), 1, 9 - nchar(cusips[short])))

  # shorten cusips longer than 9 characters
  long <- nchar(cusips) > 9
  cusips[long] <- substr(cusips[long], 1, 9)

  # update table
  table$cusipno <- cusips
  table$other <- others

  # drop rows with only letters
  table %>% 
    filter(grepl("[0-9]", cusipno)) %>% 
    mutate(check = checksum.check(cusipno)) %>% 
    select(cusipno, check) %>% 
    distinct(cusipno, .keep_all = TRUE)
}
```

# 4. Loop over all quarters in "input"

```{r}
for (i in seq_len(nrow(input))) {
  this_link  <- input$link[i]
  this_yrqtr <- input$year_qtr[i]
  this_yr    <- year(input$date[i])
  pre2004    <- this_yr < 2004

  sqlite_file <- sqlite_path(this_yrqtr)

  cat("Processing", i, "of", nrow(input), "year_qtr =", this_yrqtr, "\n")

  start_time <- Sys.time()
  message(start_time)

  cusip_df <- try(download.sec.list(this_link, this_yr), silent = TRUE)
  if (inherits(cusip_df, "try-error")) {
    cat("  → FAILED, skipping", this_yrqtr, "\n\n")
    next
  }

  con <- dbConnect(SQLite(), sqlite_file)

  if (pre2004) {
    clean_cusip <- clean.cusip.list.pre(cusip_df)
  } else {
    clean_cusip <- clean.cusip.list.post(cusip_df)
  }

  cat("  →", nrow(clean_cusip), "unique 13F CUSIPs\n\n")

  dbWriteTable(
    con,
    "cusip_list",
    clean_cusip,
    overwrite = TRUE
  )

  dbDisconnect(con)

  end_time <- Sys.time()
  elapsed <- end_time - start_time

  message(
    sprintf("Total time for %s: %.1f minutes (%.1f seconds)",
            yq, as.numeric(elapsed, units = "mins"),
            as.numeric(elapsed, units = "secs"))
}
```

# 5. Output Summary

```{r}
cusip_summary <- tibble(
  year_qtr = dates$year_qtr,
  n_cusip  = NA_integer_
)

for (i in seq_along(dates$year_qtr)) {
  yq <- dates$year_qtr[i]

  message(sprintf(
    "\n=== Quarter %d/%d: %s ===",
    i, length(dates$year_qtr), yq
  ))
  message(Sys.time())

  sqlite_file <- sqlite_path(yq)

  if (!file.exists(sqlite_file)) {
    message("No SQLite file for ", yq, "; skipping.")
    next
  }

  con <- dbConnect(SQLite(), sqlite_file)

  if (!dbExistsTable(con, "cusip_list")) {
    message("No cusip_list table for ", yq, "; skipping.")
    dbDisconnect(con)
    next
  }

  n_rows <- tbl(con, "cusip_list") %>% tally(name = "n") %>% pull(n)
  cusip_summary$n_cusip[i] <- n_rows
  dbDisconnect(con)
}

cusip_summary <- cusip_summary %>%
  mutate(
    year = as.integer(substr(year_qtr, 1, 4)),
    qtr  = as.integer(substr(year_qtr, 5, 5))
  ) %>%
  arrange(year, qtr)

p_cusip_qa <- ggplot(cusip_summary, aes(x = year + (qtr - 1) / 4, y = n_cusip)) +
  geom_line(color = "steelblue") +
  geom_point(size = 0.6) +
  scale_x_continuous(
    breaks = seq(min(cusip_summary$year), max(cusip_summary$year), by = 2),
    minor_breaks = NULL
  ) +
  labs(
    title = "Number of CUSIPs by Quarter",
    x = "Year",
    y = "Number of CUSIPs"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank()
  )

ggsave(
  filename = file.path(qa_dir_01b, "cusip_summary.png"),
  plot     = p_cusip_qa,
  width    = 8,
  height   = 4,
  dpi      = 300
)
```