---
title: "03_parse"
author: "Vinci Chan"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc_expand: true
    embed-resources: true
execute:
  warning: false
  message: false
---

# 1. Load Package

```{r}
library(tidyverse)
library(DBI)
library(RSQLite)
library(XML)
library(future)
library(furrr)

source("R/config.R")
source("R/utils_dates.R")
source("R/utils_db.R")
source("R/utils_cusip.R")
```

# 2. Setup

```{r}
dates <- make_quarter_tbl(
  start_year    = 1996,
  end_year      = cfg$end_year
)

qa_dir_03 <- file.path(cfg$qa_dir, "03")
dir.create(qa_dir_03, recursive = TRUE, showWarnings = FALSE)
```

# 3. Functions

## 3.1. Determine table type

```{r}
determine.type <- function(table, numseps = 3, minspace = 3, min.frac = 0.5) {
  table <- str_split(table, "\n") %>% pluck(1)

  # 1. blank tables 
  if (length(table) == 0) return("none")

  # 2. xml
  if (any(grepl("<[^/]*informationTable|</[^>]*informationTable", table))) return("xml")

  # 3. other table types
  n <- length(table)
  
  # compute separator scores: proportion of table rows that is separated with the delimiter
  tabseps <- sum(lengths(strsplit(table, "\t")) - 1 >= numseps) / n
  comseps <- sum(lengths(strsplit(table, ",")) - 1 >= numseps) / n
  spaceseps <- sum(
    vapply(
      gregexpr("[[:alnum:]]\\s+", table), # search how many spaces used as delimiter each line
      function(x) ifelse(all(x == -1), 0L, max(attr(x, "match.length"))), 
      integer(1)
    ) >= minspace
  ) / n

  # find winner
  scores <- c(tab = tabseps, csv = comseps, fwf = spaceseps)
  best <- which.max(scores)

  if (is.na(scores[best]) || scores[best] < min.frac) return('csv')

  return(names(best))
}
```

## 3.2. Non-XML tables

Function to quickly find the first instance of where any of "pattern" substrings occur in a string

```{r}
regmatches.multiple <- function(pattern, x, n, min.digits = 6, max.digits = 9) {
  pattern_groups <- split(pattern, ceiling(seq_along(pattern) / n))

  # find matches for each pattern group, keep row index
  match_tbl <- map_dfr(pattern_groups, function(pat_grp) {
    regex_pat <- str_c(pat_grp, collapse = "|")
    m <- str_locate(x, regex(regex_pat, ignore_case = TRUE))

    tibble(
      row   = seq_along(x),
      start = m[, "start"],
      end   = m[, "end"]
    )
  })

  # one row per original line: earliest non-NA start, matching end
  match_tbl <- match_tbl %>%
    group_by(row) %>%
    summarise(
      start = suppressWarnings(min(start, na.rm = TRUE)),
      end   = suppressWarnings(end[which.min(start)]),
      .groups = "drop"
    ) %>%
    mutate(
      start = ifelse(is.infinite(start), NA_integer_, start),
      end   = ifelse(is.infinite(start), NA_integer_, end)
    )

  # get a chunk of 6-9 digits and allow for hyphen and spaces
  cusip_regex <- str_c(
    "^([[:alnum:]][ -]*){", min.digits - 1, ",", max.digits - 1, "}[[:alnum:]]"
  )

  base_tbl <- tibble(
    row = seq_along(x),
    x   = x
  ) %>%
    left_join(match_tbl, by = "row") %>%
    mutate(
      tail = if_else(
        is.na(start),
        NA_character_,
        str_sub(x, start)
      ),
      cusip_span = if_else(
        is.na(tail),
        NA_character_,
        str_extract(tail, regex(cusip_regex))
      ),
      match = if_else(
        is.na(cusip_span),
        NA_character_,
        str_remove_all(cusip_span, "[[:space:]-]")
      ),
      match = if_else(
        !is.na(match) &
          str_length(match) >= min.digits &
          str_length(match) <= max.digits,
        match,
        NA_character_
      ),
      end = if_else(
        is.na(cusip_span),
        NA_integer_,
        start + str_length(cusip_span) - 1
      )
    )

  list(
    match = base_tbl$match,
    start = base_tbl$start,
    end   = base_tbl$end
  )
}
```

Function to split a row into cusip and its remainder, and extract character data

```{r}
cusip.remainder <- function(table, cusip_list) {
  # find cusip matches and positions
  cusip_key <- toupper(str_sub(cusip_list, 1, 6)) %>% unique() # get 6-digit issuer keys
  out <- regmatches.multiple(cusip_key, table, max.digits = 9, n = 100)
  cusip9 <- toupper(out$match)
  end <- out$end

  # shrsOrPrnAmt
  tail_str <- str_sub(table, end, nchar(table))

  shrsOrPrnAmt <- case_when(
      str_detect(tail_str, regex("SH",   ignore_case = TRUE)) ~ "SH",
      str_detect(tail_str, regex("PRN|PRIN", ignore_case = TRUE)) ~ "PRN",
      TRUE ~ NA_character_
    )

  # putCall
  putCall <- case_when(
      str_detect(tail_str, regex("PUT",   ignore_case = TRUE)) ~ "PUT",
      str_detect(tail_str, regex("CALL",  ignore_case = TRUE)) ~ "CALL",
      TRUE ~ NA_character_
    )

  # investmentDiscretion
  investmentDiscretion <- case_when(
      str_detect(tail_str, regex("SOLE",             ignore_case = TRUE)) ~ "SOLE",
      str_detect(tail_str, regex("OTHER|OTR|OTH",    ignore_case = TRUE)) ~ "OTR",
      str_detect(tail_str, regex("DEFINED|DFND|DEF", ignore_case = TRUE)) ~ "DFND",
      str_detect(tail_str, regex("SHARED",           ignore_case = TRUE)) ~ "SHARED",
      TRUE ~ NA_character_
    )

  strings_to_cut <- c(
      "SH", "PRN", "PRIN", "CALL", "PUT", "SOLE", "OTHER", "OTR", "OTH", "DEFINED", "DFND", "DEF", "SHARED", "YES"
    )

  regex_cut <- paste0(
      "(^|[^[:alpha:]]{1})(",
      paste0(strings_to_cut, collapse = "|"),
      ")([^[:alpha:]]{1}|$)"
    )

  df <- tibble(
    line = table,
    end  = end
  ) %>%
  mutate(
    tail_str = str_sub(line, end, nchar(line)),
    loc = str_locate(tail_str, regex(regex_cut, ignore_case = TRUE)),
    marker_start = loc[, "start"], # finds the start of the regex_cut match
    remainder_end = if_else(
      is.na(marker_start),
      nchar(line),
      end - 2L + marker_start # stop remainder one character earlier
    ),
    remainder = str_sub(line, end + 1, remainder_end)
  )

  tibble(
    cusip9  = cusip9,
    cusip8  = str_sub(cusip9, 1, 8),
    shrsOrPrnAmt = shrsOrPrnAmt,
    putCall = putCall,
    investmentDiscretion = investmentDiscretion,
    remainder = df$remainder
  )
}
```

Function to separate the remainder into groups of numbers

```{r}
separate.remainder <- function(rows, regex = "(\\d+|\\d{1,3}(,\\d{3})*)(\\.\\d*)?") {
  # extract all numeric substrings per row (list-column)
  numbers <- regmatches(rows, gregexpr(regex, rows))

  # build tibble: one row per original row, list-cols hold raw and numeric values
  tibble(
    idx = seq_along(numbers),
    raw = numbers
  ) %>% 
    mutate(
      # remove commas, convert to numeric; empty -> NA
      nums = lapply(raw, 
        function(x) {
          if (length(x) == 0) return(NA_real_)
          as.numeric(gsub(",", "", x))
        }
      )
    ) %>% 
      # unnest to long format: one row per number
      unnest_longer(nums, values_to = "value") %>% 
      group_by(idx) %>% 
      mutate(col_id = row_number()) %>% 
      ungroup() %>% 
      select(idx, col_id, value) %>% 
      # widen
      pivot_wider(
        names_from = col_id,
        values_from = value,
        names_prefix = "num"
      ) %>% 
      arrange(idx) %>% 
      select(-idx)
}
```

Function to parse all non-xml tables (calls `cusip.remainder()`, `separate.remainder()`)

```{r}
parse.others <- function(table, cusip_list) {
  # extract CUSIP and character field
  data <- cusip.remainder(table, cusip_list)

  # extract numeric fields
  numbers <- separate.remainder(data$remainder)

  # combined metadata + numeric columns
  data <- bind_cols(data, numbers)

  # rename num* column assuming first = value, second = shares
  num_cols <- names(data) %>% keep(~ str_starts(.x, "num"))

  if (length(num_cols) == 0L) {
    warning("No numeric columns (num*) found in parse_others.", immediate. = TRUE)
    value_col  <- NA_real_
    shares_col <- NA_real_
  } else if (length(num_cols) == 1L) {
    value_col  <- NA_real_
    shares_col <- data[[num_cols[1]]]
  } else {
    value_col  <- data[[num_cols[1]]]
    shares_col <- data[[num_cols[2]]]
  }

  tibble(
    nameOfIssuer        = NA_character_,
    titleOfClass        = NA_character_,
    cusip               = data$cusip9,     
    value               = value_col,
    shrsOrPrnAmt        = data$shrsOrPrnAmt,
    investmentDiscretion = data$investmentDiscretion,
    votingAuthority     = NA_character_,   
    shares              = shares_col
  )
}
```


## 3.3. XML tables
```{r}
parse.xml <- function(table, cusip_list) {
  info_start <- which(str_detect(table, "^\\s*<([A-Za-z0-9_]+:)?informationTable\\b"))[1]
  if (!is.na(info_start)) {
    xml_start <- info_start
  } else {
    xml_start <- which(str_detect(table, "^\\s*<\\?xml"))[1]
    if (is.na(xml_start)) xml_start <- 1L
  }

  xml_content <- table[xml_start:length(table)] %>% 
    paste(collapse = "\n")

  xml_doc <- tryCatch(
    xmlParse(xml_content),
    error = function(e) {
      warning("XML parse failed: ", conditionMessage(e), immediate. = FALSE)
      return(NULL)
    }
  )

  # if parse failed, return empty tibble
  if (is.null(xml_doc)) {
    return(tibble(
      nameOfIssuer         = character(),
      titleOfClass         = character(),
      cusip                = character(),
      value                = double(),
      shrsOrPrnAmt         = character(),
      investmentDiscretion = character(),
      votingAuthority      = character(),
      shares               = double()
    ))
  }

  xmlToDataFrame(xml_doc, stringsAsFactors = FALSE) %>% 
    as_tibble() %>% 
    mutate(
      cusip = str_remove_all(cusip, "[[:space:]-]"),
      value = as.numeric(str_remove_all(value, "\n|\t| ")),
      shares_temp = shrsOrPrnAmt,
      shrsOrPrnAmt = case_when(
        str_detect(shrsOrPrnAmt, "SH")  ~ "SH",
        str_detect(shrsOrPrnAmt, "PRN") ~ "PRN",
        TRUE ~ NA_character_
      ),
      shares = suppressWarnings(
        as.numeric(str_remove_all(shares_temp, "\n|\t|SH|PRN| "))
      ) 
    ) %>% 
    select(-shares_temp)
}
```

## 3.4. Function to parse any table

```{r}
gsub.all <- function(pattern, replacement, x) {
  if (length(pattern) == 0) return(x)

  for (i in seq_along(pattern)) {
    if (identical(pattern[i], "")) next
    x <- gsub(pattern[i], replacement[i], x, fixed = TRUE)
  }
  x
}
```

```{r}
parse.one.table <- function(raw_text, cusip_list) {
  table <- raw_text %>% str_split("\n") %>% pluck(1)
  type  <- determine.type(raw_text)

  if (type == "none") {
    return(tibble(
      nameOfIssuer         = character(),
      titleOfClass         = character(),
      cusip                = character(),
      value                = double(),
      shrsOrPrnAmt         = character(),
      investmentDiscretion = character(),
      votingAuthority      = character(),
      shares               = double()
    ))
  }

  if (type == "tab") {
    table <- str_replace_all(table, "\t", " ")
  } else if (type == "csv") {
    quoted         <- regmatches(table, gregexpr("(\")(.*?)(\")", table))
    quoted_nocomma <- map(quoted, ~ gsub("\\\"|,", "", .x))

    table <- map_chr(seq_along(table), ~ {
      gsub.all(quoted_nocomma[[.x]], quoted[[.x]], table[.x])
    })

    table <- str_replace_all(table, "\t|,", " ")
  }

  out <- if (type == "xml") {
    parse.xml(table, cusip_list)
  } else {
    parse.others(table, cusip_list)
  }

  out %>%
    mutate(table_type = type) %>%
    select(
      nameOfIssuer,
      titleOfClass,
      cusip,
      value,
      shrsOrPrnAmt,
      investmentDiscretion,
      votingAuthority,
      shares,
      table_type
    )
}
```

# 4. Quarterly wrapper

```{r}
build.quarter.tables <- function(year_qtr) {
  con <- connect_sqlite(year_qtr)
  on.exit(disconnect_sqlite(con), add = TRUE)

  header_table <- tbl(con, "header_table") %>% as_tibble()

  holdings_table_raw <- tbl(con, "holdings_table_raw") %>% as_tibble()

  if (nrow(holdings_table_raw) == 0L) {
    return(invisible(NULL))
  }

  yr  <- as.integer(substr(year_qtr, 1, 4))
  use_cusip_join <- (yr >= 1996)

  cusip_list <- if (use_cusip_join) get.cusip.vec(con, year_qtr) else character(0)

  message("Parsing ", nrow(holdings_table_raw), " filings in parallel...")

  plan(multisession)

  parsed_list <- future_map(
    as.character(holdings_table_raw$holding_text),
    ~ parse.one.table(.x, cusip_list),
    .progress = TRUE
  )

  plan(sequential)

  # add filename, parsed list-column, and raw row count
  raw_with_counts <- holdings_table_raw %>% 
    mutate(
      filename = FILENAME,
      parsed = parsed_list,
      raw_nrows = str_split(as.character(holding_text), "\n") %>% 
        map_int(~ sum(.x != ""))
    )

  parsed_q <- raw_with_counts %>%
    select(filename, parsed) %>%
    unnest(parsed) %>%
    left_join(
      header_table %>%
        select(filename = FILENAME, form, cik, cikname, rdate, fdate),
      by = "filename"
    ) %>%
    select(
      form, cik, cikname, rdate, fdate, filename,
      nameOfIssuer, titleOfClass, cusip, value,
      shrsOrPrnAmt, investmentDiscretion, votingAuthority, shares,
      table_type
    )

  dbWriteTable(
    con,
    "holdings_table_parsed",
    parsed_q,
    overwrite = TRUE
  )

  rm(
    header_table, holdings_table_raw, cusip_list,
    parsed_list, raw_with_counts, parsed_q
  )

  gc()
}
```

# 5. Loop over all quarters

```{r}
for (i in seq_along(dates$year_qtr)) {
  yq <- dates$year_qtr[i]
  message("=== Starting quarter ", yq, " (", i, "/", length(dates$year_qtr), ") ===")

  start_time <- Sys.time()
  message(start_time)

  build.quarter.tables(yq)

  end_time <- Sys.time()
  elapsed <- end_time - start_time

  message(
    sprintf("Total time for %s: %.1f minutes (%.1f seconds)",
            yq, as.numeric(elapsed, units = "mins"),
            as.numeric(elapsed, units = "secs"))
  )

  gc()
}
```

# 6. Parse summary

```{r}
parse_summary <- tibble(
  year_qtr       = dates$year_qtr,
  n_raw_nonzero  = NA_integer_,
  n_parsed_some  = NA_integer_
)

for (i in seq_along(dates$year_qtr)) {
  yq <- dates$year_qtr[i]
  message(sprintf(
    "\n=== Quarter %d/%d: %s ===",
    i, length(dates$year_qtr), yq
  ))

  con <- connect_sqlite(yq)

  raw_counts <- tbl(con, "holdings_table_raw") %>%
    group_by(FILENAME) %>%
    summarise(raw_nrows = n(), .groups = "drop")

  parsed_counts <- tbl(con, "holdings_table_parsed") %>%
    group_by(filename) %>%
    summarise(parsed_nrows = n(), .groups = "drop")

  joined <- raw_counts %>%
    left_join(parsed_counts, by = c("FILENAME" = "filename")) %>%
    collect() %>% 
    mutate(parsed_nrows = replace_na(parsed_nrows, 0L))

  parse_summary$n_raw_nonzero[i] <- sum(joined$raw_nrows > 0)
  parse_summary$n_parsed_some[i] <- sum(joined$parsed_nrows > 0)

  disconnect_sqlite(con)
}

parse_summary <- parse_summary %>%
  mutate(
    year_qtr = as.numeric(year_qtr),
    pct_parsed_raw = n_parsed_some / pmax(n_raw_nonzero, 1L),
    year = year_qtr %/% 10,
    qtr  = year_qtr %% 10,
    quarter_date = case_when(
      qtr == 1 ~ ymd(paste0(year, "-02-15")),
      qtr == 2 ~ ymd(paste0(year, "-05-15")),
      qtr == 3 ~ ymd(paste0(year, "-08-15")),
      qtr == 4 ~ ymd(paste0(year, "-11-15")),
      TRUE     ~ NA_Date_
    )
  ) %>%
  arrange(quarter_date)

p_parse_qa <- ggplot(parse_summary, aes(x = quarter_date, y = pct_parsed_raw)) +
  geom_hline(yintercept = 0.95, linetype = "dashed", color = "grey60") +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 1.5) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, 1)
  ) +
  labs(
    title = "Parsing success rate over time",
    subtitle = "Share of filings with raw tables that produced parsed rows",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

write_csv(parse_summary, file.path(qa_dir_03, "parse_summary.csv"))

ggsave(
  filename = file.path(qa_dir_03, "parse_summary.png"),
  plot     = p_parse_qa,
  width    = 8,
  height   = 4,
  dpi      = 300
)
```

# 7. Inspection

```{r}
#| eval: false

# individual quarter inspection
year_qtr <- 20093
con <- connect_sqlite(year_qtr)

raw_counts <- tbl(con, "holdings_table_raw") %>%
  group_by(FILENAME) %>%
  summarise(raw_nrows = n(), .groups = "drop")

parsed_counts <- tbl(con, "holdings_table_parsed") %>%
  group_by(filename) %>%
  summarise(parsed_nrows = n(), .groups = "drop")

filingQA <- raw_counts %>%
  left_join(parsed_counts, by = c("FILENAME" = "filename")) %>%
  mutate(parsed_nrows = replace_na(parsed_nrows, 0L))

failure <- filingQA %>%
  filter(parsed_nrows == 0 & raw_nrows != 0) %>%
  pull(FILENAME)

# look at first failure
failed_raw <- tbl(con, "holdings_table_raw") %>%
  filter(FILENAME %in% failure) %>%
  as_tibble() %>%
  mutate(raw = str_split(holding_text, "\n")) %>%
  select(-holding_text)

failed_raw %>%
  pull(raw) %>%
  pluck(1) %>%
  head(20)

failed_parsed <- tbl(con, "holdings_table_parsed") %>%
  filter(filename %in% failure) %>%
  as_tibble()

disconnect_sqlite(con)
rm(con, raw_counts, parsed_counts, filingQA, failure, failed_raw, failed_parsed)
```