---
title: "05_clean"
author: "Vinci Chan"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc_expand: true
    embed-resources: true
execute:
  warning: false
  message: false
---

# 1. Setup

```{r}
library(tidyverse)
library(arrow)
library(future)
library(furrr)

source("R/config.R")
source("R/utils_dates.R")

plan(multisession)

rquarters <- make_quarter_tbl(
  start_year = cfg$start_year,
  end_year = 2025,
  start_month_day = "0101",
  end_month_day = "1231"
) %>% 
mutate(
  rquarter = floor_date(date, "quarter") + months(3) - days(1)
  ) %>% 
filter(rquarter >= ymd(19990101)) %>% 
distinct(rquarter) %>% 
arrange(rquarter)

qa_dir_05 <- file.path(cfg$qa_dir, "05")
dir.create(qa_dir_05, recursive = TRUE, showWarnings = FALSE)
```

# 2. Functions

## 2.1. Extract rows and columns

```{r}
build_holdext_raw <- function() {
  all_path <- file.path(cfg$derived_dir, "all_holdings.parquet")
  extract_path <- file.path(cfg$derived_dir, "holdext_raw_parquet")

  if(!file.exists(extract_path)) {
    open_dataset(all_path) %>% 
    # Drop rows that are completely empty on key holding fields
    filter(
      !(
        is.na(nameOfIssuer) &
        is.na(titleOfClass) &
        is.na(cusip) &
        is.na(value) &
        is.na(shrsOrPrnAmt) &
        is.na(investmentDiscretion) &
        is.na(votingAuthority) &
        is.na(shares)
      )
    ) %>% 
    # Select only relevant rows
    select(
      filename, cik, cikname,
      rdate, fdate, nameOfIssuer,
      cusip, value, shares
    ) %>% 
    # Write holdext_raw
    write_dataset(
      extract_path,
      format = "parquet"
    )
  }
  open_dataset(extract_path) %>% 
    mutate(filename = cast(filename, string()))
}
```

## 2.2. Data Cleaning

### 2.2.1 rdate

We first clean and standardise the reporting and filing dates. `rdate` and `fdate` are converted from Unix time to calendar dates, obvious data‐entry errors in `rdate` are corrected manually, and all reports are mapped to canonical quarter‑end dates (`rquarter`) to simplify building manager-quarter panels and computing returns. To focus on the post‑EDGAR regime, we then drop all pre‑1999 reporting quarters.

```{r}
clean_dates <- function(holdext_raw) {
  # Change rdate and fdate class from Unix time to date
  holdext_raw <- holdext_raw %>% 
    mutate(
      rdate = as_date(rdate),
      fdate = as_date(fdate)
    )

  # Check for weird rdates
  weird_rdates <- holdext_raw %>% 
    to_duckdb() %>% 
    filter(rdate > fdate) %>% 
    select(filename, rdate, fdate) %>% 
    distinct(filename, .keep_all = TRUE) %>% 
    collect()

  holdext_raw <- holdext_raw %>% 
    mutate(
      # Fix weird dates manually
      rdate = case_when(
        filename == "0000080255-03-000400.txt" ~ as_date("2002-12-31"),
        filename == "0000905148-07-001531.txt" ~ as_date("2006-09-30"),
        filename == "0001103882-11-000001.txt" ~ as_date("2010-09-30"),
        TRUE ~ rdate
      ),
      # for filings with rdate == fdate, assume rdate to be the previous quarter end
      same_date = rdate == fdate,
      rdate = if_else(
        same_date,
        as_date(floor_date(fdate, unit = "quarter") - ddays(1)),
        rdate
      )
    ) %>%
    select(-same_date)

  # Standardise to rquarter and handle rdate == fdate
  rquarter_build <- holdext_raw %>% 
    distinct(rdate) %>% 
    collect() %>% 
    mutate(
      rquarter = floor_date(rdate, "quarter") + months(3) - days(1)
    )

  holdext_raw %>% 
    left_join(rquarter_build, by = "rdate") %>%
    relocate(rquarter, .after = rdate) %>%
    # Remove pre-1999 rquarters
    filter(rquarter >= ymd(19990101))
}
```

Weird `rdate`:
[0000080255-03-000400.txt](https://www.sec.gov/Archives/edgar/data/80255/0000080255-03-000400.txt)
[0000905148-07-001531.txt](https://www.sec.gov/Archives/edgar/data/1352851/0000905148-07-001531.txt)
[0001103882-11-000001.txt](https://www.sec.gov/Archives/edgar/data/1352910/0001103882-11-000001.txt)

### 2.2.2. fdate

Some rdates map to the same rquarter (quarter-end bucket), with multiple filings per manager-quarter. Some subsequent filings restate the complete, updated portfolio, while some are small additions that do not represent the full portfolio.

To obtain a single canonical portfolio per manager-quarter, for each `cik` + `rquarter` we first identify the filing(s) with the maximum number of holdings (n_holdings), as these are most likely to be full portfolio snapshots. If several filings have the same maximum n_holdings, we then select the one with the latest fdate as the final version for that reporting period. All non‑selected filings are dropped.

```{r}
select_canonical_filings <- function(holdext_raw) {
  canonical <- holdext_raw %>% 
  to_duckdb() %>% 
  group_by(cik, rquarter, filename) %>% 
  summarise(
    n_holdings = n(),
    fdate = max(fdate),
    .groups = "drop"
  ) %>% 
  group_by(cik, rquarter) %>% 
  filter(n_holdings == max(n_holdings)) %>% 
  slice_max(fdate, n = 1, with_ties = FALSE) %>% 
  ungroup() %>% 
  select(cik, rquarter, filename) %>% 
  collect()

  holdext_raw %>% 
    semi_join(canonical, by = join_by(cik, rquarter, filename))
}
```

### 2.2.3. cusip 

A few thousand of rows has cusips shorter than 7 characters. Those will be dropped.

```{r}
clean_cusip <- function(holdext_raw) {
  holdext_raw %>% 
    mutate(
      cusip = toupper(str_trim(cusip)),
      cusip_len = nchar(cusip)
    ) %>% 
    filter(cusip_len >= 7) %>% 
    select(-cusip_len)
}
```

There a few thousand rows with cusip as "000000000". They are kept for now.

### 2.2.4. value and shares outliers

```{r}
filter_outliers <- function(holdext_raw) {  
  caps <- holdext_raw %>%
    to_duckdb() %>% 
    summarise(
      value_cap = quantile(value, 0.999, na.rm = TRUE),
      shares_cap = quantile(shares, 0.999, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    collect()

  holdext_raw %>%
    filter(
      (is.na(value) | value <= caps$value_cap) & 
      (is.na(shares) | shares <= caps$shares_cap)
    )
}
```

### 2.2.5. value unit conversion

Starting on January 3, 2023, market value is reported in dollars. Previously, market value was reported in thousands. [SEC Notice](https://www.sec.gov/files/form_13f_readme.pdf)

To preserve consistency over time, values reported in dollars should be converted back to thousands. However, not all filers appear to switch units at the same time, so we detect likely unit-switch episodes from manager-level quarter-on-quarter jumps in aggregate portfolio value rather than applying a blanket conversion rule.

```{r}
detect_unit_switches <- function(holdext_raw) {
  conversion_date <- ymd(20230103)

  # 1. Create manager–quarter value panel
  value_panel <- holdext_raw %>% 
    to_duckdb() %>% 
    group_by(cik, rquarter) %>% 
    summarise(
      fdate           = max(fdate, na.rm = TRUE),
      portfolio_value = sum(value,  na.rm = TRUE),
      holdings_count  = n(),
      total_shares    = sum(shares, na.rm = TRUE),
      .groups         = "drop"
    ) %>% 
    collect() %>% 
    arrange(cik, rquarter) %>% 
    group_by(cik) %>% 
    mutate(
      lag_value          = lag(portfolio_value),
      lag_holdings_count = lag(holdings_count),
      lag_total_shares   = lag(total_shares),

      value_qoq_ratio = if_else(
        portfolio_value > 0 & lag_value > 0,
        portfolio_value / lag_value,
        NA_real_
      ),
      holdings_qoq_ratio = if_else(
        holdings_count > 0 & lag_holdings_count > 0,
        holdings_count / lag_holdings_count,
        NA_real_
      ),
      shares_qoq_ratio = if_else(
        total_shares > 0 & lag_total_shares > 0,
        total_shares / lag_total_shares,
        NA_real_
      ),
      lead_value_qoq_ratio = lead(value_qoq_ratio),
      min_rquarter_per_cik = min(rquarter, na.rm = TRUE)
    ) %>% 
    ungroup()

  # 2. Calibrate thresholds using pre‑2023 data
  # Start with very loose provisional mins so we can measure the distribution
  qoq_pre2023_raw <- value_panel %>%
    filter(fdate < conversion_date)

  # Helper for levels
  calibrate_level <- function(df, var) {
    df %>%
      transmute(x = {{ var }}) %>%
      filter(!is.na(x), is.finite(x), x > 0) %>%
      summarise(
        p01  = quantile(x, 0.01,  na.rm = TRUE),
        p05  = quantile(x, 0.05,  na.rm = TRUE),
        p10  = quantile(x, 0.10,  na.rm = TRUE),
        med  = median(x, na.rm = TRUE),
        .groups = "drop"
      )
  }

  level_value_calibration    <- calibrate_level(qoq_pre2023_raw, portfolio_value)
  level_holdings_calibration <- calibrate_level(qoq_pre2023_raw, holdings_count)

  # Data‑driven minimums: e.g. 1st percentile for value and 5th percentile for holdings
  min_portfolio_value <- as.numeric(level_value_calibration$p01)
  min_holdings_count  <- as.numeric(level_holdings_calibration$p05)

  # Now filter using these calibrated mins
  qoq_pre2023 <- value_panel %>%
    filter(
      fdate < conversion_date,
      portfolio_value    >= min_portfolio_value,
      lag_value          >= min_portfolio_value,
      holdings_count     >= min_holdings_count,
      lag_holdings_count >= min_holdings_count
    )

  # Ratio calibration helper (unchanged)
  calibrate_ratio <- function(df, var) {
    df %>%
      transmute(x = {{ var }}) %>%
      filter(!is.na(x), is.finite(x), x > 0) %>%
      summarise(
        p99   = quantile(x, 0.99,  na.rm = TRUE),
        p995  = quantile(x, 0.995, na.rm = TRUE),
        p999  = quantile(x, 0.999, na.rm = TRUE),
        max_x = max(x, na.rm = TRUE),
        .groups = "drop"
      )
  }

  value_calibration    <- calibrate_ratio(qoq_pre2023, value_qoq_ratio)
  holdings_calibration <- calibrate_ratio(qoq_pre2023, holdings_qoq_ratio)
  shares_calibration   <- calibrate_ratio(qoq_pre2023, shares_qoq_ratio)

  # Threshold for unit_conversions
  jump_on_threshold  <- min(250, value_calibration$p995)
  jump_off_threshold <- 1 / jump_on_threshold

  # Post‑jump stabilisation (value ratios)
  stable_lower <- 0.67
  stable_upper <- 1.50

  # Holdings and shares jump thresholds (ratio caps, data‑driven)
  holdings_jump_upper <- min(5, holdings_calibration$p999)
  shares_jump_upper   <- min(10, shares_calibration$p99 * 2)

  # 3. Detect likely switch episodes
  value_flags <- value_panel %>% 
    filter(fdate >= conversion_date) %>% 
    arrange(cik, rquarter) %>% 
    group_by(cik) %>% 
    mutate(
      has_preconversion_data = min_rquarter_per_cik < conversion_date,

      eligible_for_detection =
        portfolio_value    >= min_portfolio_value &
        lag_value          >= min_portfolio_value &
        holdings_count     >= min_holdings_count &
        lag_holdings_count >= min_holdings_count,

      stable_next_qtr =
        !is.na(lead_value_qoq_ratio) &
        dplyr::between(lead_value_qoq_ratio, stable_lower, stable_upper),

      holdings_stable =
        !is.na(holdings_qoq_ratio) &
        holdings_qoq_ratio <= holdings_jump_upper,

      shares_stable =
        !is.na(shares_qoq_ratio) &
        shares_qoq_ratio <= shares_jump_upper,

      switch_on_event =
        !is.na(value_qoq_ratio) &
        eligible_for_detection &
        holdings_stable &
        shares_stable &
        value_qoq_ratio >= jump_on_threshold &
        stable_next_qtr,

      switch_off_event =
        !is.na(value_qoq_ratio) &
        eligible_for_detection &
        holdings_stable &
        shares_stable &
        value_qoq_ratio <= jump_off_threshold,

      event          = dplyr::case_when(
        switch_on_event  ~ 1L,
        switch_off_event ~ -1L,
        TRUE             ~ 0L
      ),
      raw_episode     = cumsum(event),
      active_episode  = pmax(raw_episode, 0L),
      # Apply scaling: detected switch OR post-conversion-only fund
      scale_down_flag = active_episode > 0 | !has_preconversion_data
    ) %>% 
    ungroup()

  # 4. Apply scaling to full row‑level data
  holdext_raw %>% 
    left_join(
      value_flags %>% 
        select(cik, rquarter, scale_down_flag),
      by = c("cik", "rquarter"),
      copy = TRUE
    ) %>% 
    mutate(
      scale_down_flag = if_else(is.na(scale_down_flag), FALSE, scale_down_flag),
      value = if_else(scale_down_flag, value / 1000, value)
    ) %>%
    select(
      filename, cik, cikname, rdate, rquarter, fdate,
      nameOfIssuer, cusip, value, shares
    )
}
```

### 2.2.6. Missing quarters

We build a complete cik-quarter panel with LOCF (last observation carried forward).

```{r}
build_missing_quarters_batch <- function(batch_ds) {
  # Identify the columns that define a holding row
  holding_cols <- c(
    "nameOfIssuer",
    "cusip",
    "value",
    "shares"
  )

  # 1. Load into memory
  batch_df <- batch_ds %>% collect()

  # 2. Complete grid
  cik_ranges <- batch_df %>% 
    distinct(cik, rquarter) %>% 
    group_by(cik) %>% 
    summarise(
      min_rquarter = min(rquarter),
      max_rquarter = max(rquarter),
      .groups = "drop"
    )
  
  cik_quarter_grid <- cik_ranges %>%
    cross_join(rquarters) %>%
    filter(
      rquarter >= min_rquarter,
      rquarter <= max_rquarter
    ) %>%
    select(cik, rquarter)

  filled_holdext_rquarter <- cik_quarter_grid %>% 
    left_join(batch_df, by = c("cik", "rquarter"))
  
  # 3. Flag missing vs non-operating quarters
  filled_holdext_rquarter <- filled_holdext_rquarter %>%
    mutate(
      is_empty_row = if_all(all_of(holding_cols), ~ is.na(.x)),
      missing_quarter = if_else(is_empty_row, 1L, 0L)
    ) %>%
    select(-is_empty_row)

  # 4. Compute prior_rdate for every row
  filled_with_prior <- filled_holdext_rquarter %>% 
    arrange(cik, rquarter) %>% 
    group_by(cik) %>% 
    mutate(
      own_rquarter = if_else(
        missing_quarter == 0L,
        rquarter,
        as.Date(NA)
      ),
      prior_rquarter_num = cummax(coalesce(as.numeric(own_rquarter), -Inf)),
      prior_rquarter = as_date(prior_rquarter_num)
    ) %>% 
    ungroup() %>% 
    select(-own_rquarter, -prior_rquarter_num)

  # 5. Build holdext_clean: keep real + non-OP rows, LOCF for missing
  real_holdings <- filled_with_prior %>% 
    filter(missing_quarter == 0L) %>% 
    select(cik:shares)

  core_cols <- colnames(real_holdings)

  locf_rows <- filled_with_prior %>% 
    filter(missing_quarter == 1L) %>% 
    select(cik, rquarter, prior_rquarter) %>% 
    left_join(
      real_holdings %>% 
        rename(prior_rquarter = rquarter),
      by = c("cik", "prior_rquarter"),
      relationship = "many-to-many"
    ) %>% 
    select(all_of(core_cols))

  bind_rows(real_holdings, locf_rows)
}
```

# 3. Orchestration

```{r}
# Stage 0
stage0 <- build_holdext_raw()

# Stage 1
stage1 <- stage0 %>%
  clean_dates() %>%
  select_canonical_filings() %>%
  clean_cusip() %>%
  filter_outliers() %>%
  detect_unit_switches()

# Stage 2
stage2_path <- file.path(cfg$staging_dir, "holdext_raw_stage2_parquet")
if (dir.exists(stage2_path)) unlink(stage2_path, recursive = TRUE, force = TRUE)

cik_batch <- stage1 %>%
  distinct(cik) %>%
  arrange(cik) %>%
  collect() %>%
  mutate(batch_id = sprintf("batch_%04d", ceiling(row_number() / 500)))

stage1 %>%
  left_join(cik_batch, by = "cik") %>%
  write_dataset(
    path = stage2_path,
    format = "parquet",
    partitioning = "batch_id"
  )

stage2 <- open_dataset(stage2_path)

# Stage 3
stage3_path <- file.path(cfg$derived_dir, "holdext_clean_parquet")
if (dir.exists(stage3_path)) unlink(stage3_path, recursive = TRUE, force = TRUE)
dir.create(stage3_path, recursive = TRUE)

batch_ids <- cik_batch %>%
  distinct(batch_id) %>%
  arrange(batch_id) %>%
  pull(batch_id)

future_walk(batch_ids, function(b) {
  batch_ds <- open_dataset(stage2_path) %>%
    filter(batch_id == b) %>%
    select(-batch_id)

  holdext_clean_batch <- build_missing_quarters_batch(batch_ds)

  write_parquet(
    holdext_clean_batch,
    sink = file.path(stage3_path, paste0(b, ".parquet")),
    compression = "zstd"
  )

  rm(holdext_clean_batch)
  gc()
  
  invisible(NULL)
})

holdext_clean <- open_dataset(stage3_path)
```

# 4. Check and clean up

```{r}
n_gaps <- holdext_clean %>%
  distinct(cik, rquarter) %>%
  collect() %>% 
  arrange(cik, rquarter) %>%
  group_by(cik) %>%
  mutate(
    expected_next = ceiling_date(rquarter, "month") %m+% months(3) - days(1),
    actual_next   = lead(rquarter),
    has_gap       = !is.na(actual_next) & actual_next != expected_next
  ) %>%
  ungroup() %>%
  filter(has_gap) %>%
  summarise(n_gap_rows = n()) %>%
  pull(n_gap_rows)

if(n_gaps == 0) {
  unlink(stage2_path, recursive = TRUE, force = TRUE)
}
```

# 5. QA

```{r}
holdext_raw   <- open_dataset(file.path(cfg$derived_dir,  "holdext_raw_parquet"))
holdext_clean <- open_dataset(file.path(cfg$derived_dir,  "holdext_clean_parquet"))
```

```{r}
quarter_cmp <- holdext_raw %>%
  select(cik, rdate) %>%
  collect() %>%
  mutate(
    rdate = as_date(rdate),
    rquarter = floor_date(rdate, "quarter") + months(3) - days(1)
  ) %>%
  distinct(cik, rquarter) %>%
  count(cik, name = "raw_n_q") %>%
  full_join(
    holdext_clean %>%
      distinct(cik, rquarter) %>%
      collect() %>%
      count(cik, name = "clean_n_q"),
    by = "cik"
  ) %>%
  mutate(
    raw_n_q   = replace_na(raw_n_q, 0L),
    clean_n_q = replace_na(clean_n_q, 0L),
    diff_q    = clean_n_q - raw_n_q
  )

p_quarters <- ggplot(quarter_cmp, aes(x = raw_n_q, y = clean_n_q)) +
  geom_abline(slope = 1, intercept = 0, color = "grey60", linetype = "dashed") +
  geom_point(aes(color = diff_q), alpha = 0.6, size = 1.2) +
  scale_color_gradient2(
    low = "#3b4cc0",
    mid = "#b0b0b0",
    high = "#f95d6a",
    midpoint = 0
  ) +
  labs(
    title = "Number of rquarters per CIK: holdext_raw vs holdext_clean",
    subtitle = "Points on the dashed line have equal quarter counts; points above it gained quarters in holdext_clean",
    x = "Raw number of rquarters",
    y = "Clean number of rquarters",
    color = "Clean - raw"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.margin = margin(t = 15, r = 15, b = 20, l = 15),
    plot.title.position = "plot",
    plot.subtitle = element_text(margin = margin(b = 10))
  )

ggsave(
  filename = file.path(qa_dir_05, "qa_quarters_raw_vs_clean.png"),
  plot     = p_quarters,
  width    = 8,
  height   = 5,
  dpi      = 300
)
```

```{r}
# Map raw rdate to quarter-end
raw_quarter_map <- holdext_raw %>%
  mutate(rdate = as_date(rdate)) %>%
  distinct(cik, rdate) %>%
  collect() %>% 
  mutate(
    rquarter = floor_date(rdate, "quarter") + months(3) - days(1)
  )

example_cik <- "0000002230"

raw_series <- holdext_raw %>%
  mutate(rdate = as_date(rdate)) %>%
  inner_join(raw_quarter_map, by = c("cik", "rdate")) %>%
  filter(cik == example_cik) %>%
  group_by(cik, rquarter) %>%
  summarise(raw_portfolio_value = sum(value, na.rm = TRUE), .groups = "drop")

clean_series <- holdext_clean %>%
  filter(cik == example_cik) %>%
  group_by(cik, rquarter) %>%
  summarise(clean_portfolio_value = sum(value, na.rm = TRUE), .groups = "drop")

both <- raw_series %>%
  inner_join(clean_series, by = c("cik", "rquarter")) %>%
  collect() %>%  
  mutate(
    raw_b   = raw_portfolio_value   / 1e9,
    clean_b = clean_portfolio_value / 1e9
  )

p_value <- ggplot(both, aes(x = rquarter)) +
  geom_line(aes(y = raw_b,   color = "Raw")) +
  geom_point(aes(y = raw_b,  color = "Raw"),  size = 1.2) +
  geom_line(aes(y = clean_b, color = "Clean")) +
  geom_point(aes(y = clean_b, color = "Clean"), size = 1.2) +
  labs(
    title = paste("Raw vs clean portfolio value for CIK", unique(both$cik)),
    x = NULL,
    y = "Portfolio value (billions USD)",
    color = NULL
  ) +
  theme_minimal(base_size = 12)

ggsave(
  filename = file.path(qa_dir_05, "qa_value_raw_vs_clean_example_cik.png"),
  plot     = p_value,
  width    = 7,
  height   = 4.5,
  dpi      = 300
)
```