---
title: "06_analysis"
author: "Vinci Chan"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc_expand: true
    embed-resources: true
execute:
  warning: false
  message: false
---

# 1. Setup

```{r}
library(tidyverse)
library(arrow)

source("R/config.R")
source("R/utils_dates.R")
```

# 2. Data Import

```{r}
holdext_clean <- open_dataset(file.path(cfg$derived_dir, "holdext_clean_parquet"))
```

# 3. Functions

## 3.1. Manager Selection

```{r}
calculate_fund_returns <- function(holdings_data) {
  holdings_data %>%
    to_duckdb() %>%
    group_by(cik, rquarter) %>%
    summarise(
      total_value = sum(value, na.rm = TRUE),
      n_holdings = n(),
      .groups = "drop"
    ) %>%
    collect() %>%
    arrange(cik, rquarter) %>%
    group_by(cik) %>%
    mutate(
      prev_value = lag(total_value),
      quarterly_return = (total_value - prev_value) / prev_value
    ) %>%
    ungroup()
}

score_funds_by_horizon <- function(returns_data, horizon_quarters, latest_qtr) {
  returns_data %>%
    filter(!is.na(quarterly_return)) %>%
    mutate(quarters_ago = quarter_diff(rquarter, latest_qtr)) %>%
    filter(quarters_ago <= horizon_quarters) %>%
    group_by(cik) %>%
    filter(n() == horizon_quarters) %>%
    summarise(
      n_period = n(),
      mean_return = mean(quarterly_return),
      sd_return = sd(quarterly_return),
      sharpe_ratio = if_else(sd_return == 0, NA_real_, mean_return / sd_return),
      .groups = "drop"
    )
}

assign_size_buckets <- function(returns_data, latest_qtr, n_buckets = 4) {
  returns_data %>%
    filter(rquarter == latest_qtr) %>%
    distinct(cik, total_value) %>%
    mutate(
      size_bucket = ntile(total_value, n_buckets),
      size_bucket = paste0("Q", size_bucket)
    ) %>%
    select(cik, total_value, size_bucket)
}

select_cohort_funds <- function(scored_funds, n_select = 10) {
  top_tbl <- scored_funds %>%
    arrange(desc(sharpe_ratio), cik) %>%
    slice_head(n = n_select) %>%
    mutate(cohort = "top")

  bottom_tbl <- scored_funds %>%
    arrange(sharpe_ratio, cik) %>%
    slice_head(n = n_select) %>%
    mutate(cohort = "bottom")

  mid_start <- floor((nrow(scored_funds) - n_select) / 2) + 1

  middle_tbl <- scored_funds %>%
    arrange(sharpe_ratio, cik) %>%
    slice(mid_start:(mid_start + n_select - 1)) %>%
    mutate(cohort = "middle")

  bind_rows(top_tbl, bottom_tbl, middle_tbl)
}
```

## 3.2. Portfolio Similarity

```{r}
get_holdings_at_period_start <- function(holdings_data, selected_funds_tbl) {
  holdings_data %>%
    to_duckdb() %>%
    select(cik, rquarter, cusip, value) %>%
    inner_join(
      selected_funds_tbl %>%
        select(horizon_years, start_date, size_bucket, cohort, cik),
      join_by(cik, rquarter == start_date),
      copy = TRUE
    ) %>%
    collect()
}

compute_portfolio_weights <- function(holdings_data) {
  holdings_data %>%
    group_by(horizon_years, size_bucket, cohort, cik) %>%
    mutate(
      total_value = sum(value, na.rm = TRUE),
      value_weight = value / total_value
    ) %>%
    select(horizon_years, size_bucket, cohort, cik, cusip, value_weight) %>%
    ungroup() %>%
    group_by(horizon_years, size_bucket, cohort, cik, cusip) %>%
    summarise(
      value_weight = sum(value_weight, na.rm = TRUE), # for cusips with multiple entries
      .groups = "drop"
    )
}

portfolio_vectors_to_matrices <- function(portfolio_vectors) {
  portfolio_vectors %>%
    group_by(horizon_years, size_bucket, cohort) %>%
    nest() %>%
    mutate(
      # 1. reshape to wide format
      wide = map(data, ~
        pivot_wider(
          .x,
          names_from = cusip,
          values_from = value_weight,
          values_fill = 0
        )
      ),
      # 2. turn each wide table into a matrix
      sim_matrix = map(wide, ~ {
        mat <- select(.x, -cik) %>% as.matrix()
        rownames(mat) <- .x$cik
        mat
      })
    ) %>%
    select(horizon_years, size_bucket, cohort, sim_matrix)
}

cosine_similarity_from_matrix <- function(M) {
  # M is a matrix where each ROW is one fund's portfolio (weights across CUSIPs)
  # Returns: symmetric matrix of cosine similarities between all fund pairs

  # Compute L2 norm for each row
  row_norms <- sqrt(rowSums(M * M))

  # Normalize rows to unit length
  M_norm <- M / row_norms

  # Pairwise dot products = cosine similarities
  M_norm %*% t(M_norm)
}

extract_unique_pairs <- function(sim_matrix) {
  # Convert similarity matrix to tibble and remove self-pairs and duplicates
  as_tibble(as.table(sim_matrix), .name_repair = "minimal") %>%
    set_names(c("fund_1", "fund_2", "cos_sim")) %>%
    filter(fund_1 != fund_2) %>%
    mutate(
      pair_key = if_else(
        fund_1 < fund_2,
        paste(fund_1, fund_2, sep = "_"),
        paste(fund_2, fund_1, sep = "_")
      )
    ) %>%
    group_by(pair_key) %>%
    summarise(
      fund_1 = first(fund_1),
      fund_2 = first(fund_2),
      cos_sim = mean(cos_sim),
      .groups = "drop"
    )
}

get_extreme_pairs_by_cohort <- function(cosine_sims, time_horizons) {
  cosine_sims %>%
    mutate(
      pair_tbl = map(sim_matrix, ~ {
        pairs <- extract_unique_pairs(.x)
        most_similar <- pairs %>%
          arrange(desc(cos_sim)) %>%
          slice_head(n = 1) %>%
          mutate(type = "most")
        least_similar <- pairs %>%
          arrange(cos_sim) %>%
          slice_head(n = 1) %>%
          mutate(type = "least")
        bind_rows(most_similar, least_similar)
      })
    ) %>%
    select(horizon_years, size_bucket, cohort, pair_tbl) %>%
    unnest(pair_tbl) %>%
    rename(horizon = horizon_years) %>%
    left_join(
      time_horizons %>%
        select(horizon_years, start_date),
      by = c("horizon" = "horizon_years")
    )
}
```

## 3.3. Backtest and Performance

```{r}
prepare_extreme_pairs <- function(cosine_sims, time_horizons) {
  get_extreme_pairs_by_cohort(cosine_sims, time_horizons)
}

get_pair_holdings_over_time <- function(holdings_data, pairs_table) {
  pairs_table %>%
    pivot_longer(
      cols = c(fund_1, fund_2),
      names_to = "fund_slot",
      values_to = "fund"
    ) %>%
    left_join(
      holdings_data %>%
        to_duckdb() %>%
        select(cik, rquarter, cusip, value, shares),
      join_by(fund == cik),
      relationship = "many-to-many",
      copy = TRUE
    ) %>%
    filter(rquarter >= start_date) %>%
    collect()
}

calculate_backtest_returns <- function(backtest_holdings) {
  backtest_holdings %>%
    # 1) Combine the two funds into one pair portfolio per quarter
    group_by(
      horizon, size_bucket, cohort, type,
      pair_key, rquarter
    ) %>%
    summarise(
      total_value_pair = sum(value, na.rm = TRUE),
      n_holdings_pair  = n(),
      .groups = "drop"
    ) %>%
    # 2) Order over time within each pair
    arrange(
      horizon, size_bucket, cohort, type,
      pair_key, rquarter
    ) %>%
    # 3) Compute time-series returns at the pair level
    group_by(
      horizon, size_bucket, cohort, type,
      pair_key
    ) %>%
    mutate(
      prev_value_pair   = lag(total_value_pair),
      quarterly_return  = (total_value_pair - prev_value_pair) / prev_value_pair
    ) %>%
    ungroup()
}

compute_backtest_statistics <- function(backtest_returns) {
  backtest_returns %>%
    filter(!is.na(quarterly_return)) %>%
    group_by(
      horizon, size_bucket, cohort, type,
      pair_key
    ) %>%
    summarise(
      n_period     = n(),
      mean_return  = mean(quarterly_return),
      sd_return    = sd(quarterly_return),
      sharpe_ratio = if_else(sd_return == 0, NA_real_, mean_return / sd_return),
      .groups = "drop"
    ) %>%
    arrange(horizon, size_bucket, cohort, type, desc(sharpe_ratio))
}
```

# 4. Orchestration

## 4.1. Manager Selection

```{r}
# Calculate quarterly returns for all funds
returns_by_fund <- calculate_fund_returns(holdext_clean)

# Define time horizons and identify top performers
latest_rquarter <- max(returns_by_fund$rquarter)
n_select <- 10
n_buckets <- 4

time_horizons <- tribble(
  ~horizon_years,
  2,
  3,
  4,
  5,
) %>%
  mutate(
    quarters_back = horizon_years * 4,
    start_date = latest_rquarter - months(quarters_back * 3)
  )
time_horizons

fund_sizes <- assign_size_buckets(
  returns_data = returns_by_fund,
  latest_qtr = latest_rquarter,
  n_buckets = n_buckets
)
fund_sizes

fund_cohorts <- time_horizons %>%
  mutate(
    scored_funds = map(
      quarters_back,
      ~ score_funds_by_horizon(returns_by_fund, .x, latest_rquarter)
    )
  ) %>%
  select(horizon_years, start_date, scored_funds) %>%
  unnest(scored_funds) %>%
  inner_join(fund_sizes, by = "cik") %>%
  group_by(horizon_years, start_date, size_bucket) %>%
  group_modify(~ select_cohort_funds(.x, n_select = n_select)) %>%
  ungroup()

time_horizons %>%
  mutate(
    scored_funds = map(
      quarters_back,
      ~ score_funds_by_horizon(returns_by_fund, .x, latest_rquarter)
    )
  )
fund_cohorts %>% glimpse()
```

## 4.2. Portfolio Similarity

```{r}
# Get holdings at start of each time horizon
holdings_at_start <- get_holdings_at_period_start(holdext_clean, fund_cohorts)

# Compute portfolio weights
portfolio_vectors <- compute_portfolio_weights(holdings_at_start)
portfolio_vectors

# Convert to matrices (one per horizon × size_bucket × cohort)
portfolio_matrices <- portfolio_vectors_to_matrices(portfolio_vectors)
portfolio_matrices
portfolio_matrices %>% ungroup() %>% slice(3) %>% pluck("sim_matrix") %>% glimpse() # example

# Compute cosine similarities
cosine_sims <- portfolio_matrices %>%
  mutate(sim_matrix = map(sim_matrix, cosine_similarity_from_matrix))
cosine_sims
cosine_sims %>% ungroup() %>% slice(3) %>% pluck("sim_matrix") # example

# Extract extreme pairs per horizon/size_bucket/cohort
extreme_pairs <- prepare_extreme_pairs(cosine_sims, time_horizons)
extreme_pairs
```

## 4.3. Backtest Performance

```{r}
# Get holdings for paired funds over time
backtest_holdings <- get_pair_holdings_over_time(holdext_clean, extreme_pairs)

# Calculate returns
backtest_returns <- calculate_backtest_returns(backtest_holdings)

# Compute summary statistics
backtest_stats <- compute_backtest_statistics(backtest_returns)
backtest_stats
```

# 5. Results & Interpretation

```{r}
sharpe_wide <- backtest_stats %>% 
  select(horizon, size_bucket, cohort, type, pair_key, sharpe_ratio) %>% 
  pivot_wider(
    names_from = type,
    values_from = c(sharpe_ratio, pair_key),
    names_sep = "_"
  ) %>% 
  mutate(
    sharpe_diff = sharpe_ratio_most - sharpe_ratio_least
  )
sharpe_wide

sharpe_wide %>%
  ggplot(aes(x = factor(horizon), y = sharpe_diff)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_boxplot(fill = "steelblue", alpha = 0.6, outlier.alpha = 0.6) +
  labs(
    x = "Horizon (years)",
    y = "Sharpe(most) - Sharpe(least)",
    title = "Difference in Sharpe between most- and least-similar pairs\nby horizon (within size bucket and cohort)"
  ) +
  theme_minimal()

overall <- sharpe_wide %>%
  summarise(
    n_cells        = n(),
    n_most_better  = sum(sharpe_diff > 0, na.rm = TRUE),
    n_least_better = sum(sharpe_diff < 0, na.rm = TRUE),
    mean_diff      = mean(sharpe_diff, na.rm = TRUE),
    median_diff    = median(sharpe_diff, na.rm = TRUE)
  )
overall

by_horizon <- sharpe_wide %>%
  group_by(horizon) %>%
  summarise(
    n_cells       = n(),
    n_most_better = sum(sharpe_diff > 0, na.rm = TRUE),
    mean_diff     = mean(sharpe_diff, na.rm = TRUE),
    .groups = "drop"
  )
by_horizon
```

Controlling for fund size and ex-ante performance cohort, there is no strong, consistent relationship between portfolio similarity and subsequent pair-level Sharpe ratios. Across all horizon–size–cohort cells, most-similar pairs outperform in 26 of 48 cases, with an average Sharpe difference of about 0.01–0.02, which is small. The sign of the effect varies by horizon: over shorter (2-year) windows, less-similar pairs actually do slightly better, while over intermediate (3–4 year) windows, more-similar pairs tend to show somewhat higher Sharpe. Over the longest (5-year) horizon, similarity appears to have no meaningful impact.